<style>
 td {
  width: 150px;
}
</style>
---
title: 95-868 Homework 1
author: Samantha Misurda
output: html_document
---

#### Instructions 

Submit this Rmd file on blackboard. Don't submit additional files.

Fill in the missing code chunks to answer the following questions. You should use a reasonable coding style and comment your code so that it is clear what you are doing. As a rough rule of thumb, expect to add at least one comment to each missing code chunk.

If the question asks for a numerical answer, such as "what fraction of households in the data set satisfy condition X," do not hard code the value from the computation. If the data set changes, your answer should  update automatically. 

Here are two recommended ways that you can report your answer:

**Example Question 1:** In the `cars` data set, what is the average value of the `speed` column?

```{r}
# take average of mpg$speed, save as variable
avg.speed = mean(cars$speed)
```

**Answer:** The average value was `r avg.speed`

**Example Question 1:** In the `ChickWeight` dataset, what was the average weight of the chickens grouped by each diet type?

```{r}
library(plyr)
library(knitr) # for the kable() function, which prints data frames as tables:

# group the chickens by diet and take average weight
avg.weights = ddply(ChickWeight, "Diet", summarize, avg.weight = mean(weight))
# display the data frame as an easy-to-read table, with 3 significant digits
kable(avg.weights, digits = 3, col.names = c('Diet', 'Average Weight'))
```


#### Questions 

**Note:** We will assume that the data file `dataset_hw1.csv` is in your working directory.

**Problem 1:** 

Load the CSV file `dataset_hw1.csv` into memory, using the command `read.file()` or `read.csv()` as described in lecture. Save the result as a variable called `expenditures`. 

Hint: You may want to look at the CSV file using another program such as excel to see if the first row is a header. 

```{r}
# Read in the dataset, using the header parameter to indicate that headers are present
expenditures <- read.csv("dataset_hw1.csv", header=TRUE, sep=",")
```

**Problem 2:**

a) How many records are in the data set? 

b) The column `home.type` describes the type of home belonging to each respondent. What are the different levels that this column is allowed to take?
```{r}
# The number of rows in the dataset, where each row is a record
number.of.records <- length(expenditures$housing)

# We might also want to know how many complete rows we have -- those with no missing records
number.of.complete.records <- sum(complete.cases(expenditures))

# The values of home type
home.types <- levels(expenditures$home.type)

```

**ANSWER:** 

a) There are `r number.of.records` rows in the data set.

b) The allowable home types are: `r home.types`

**Problem 3:**

What fraction of records in the data set belong to each `home.type`?

```{r}
# We can use the total number of records variable from earlier to calculate each percentage.
# First, group the data by home type, count the number of each home type, and then divide by the total number of records
proportions.by.home.type <- ddply(expenditures, "home.type", summarize, percentage.of.home.types = length(home.type)/number.of.records)

# Print out the answer in a formatted table
kable(proportions.by.home.type, digits = 3, col.names = c('Home Type', 'Proportion of Records'))
```

**Problem 4:**

Change the `expenditures` data frame so that `home.type` column has only 3 coded types: 

1. single family detached homes
2. apartments or flats
3. other (everything else)

Print the number of records in each `home.type` as a table.

```{r}

# Create the three categories
expenditures$home.type <- mapvalues(expenditures$home.type,
                          from = c("3-plex or 4-plex", "apartment or flat", "dormitory", "duplex", "end row or end townhouse", "garden", "high-rise", "mobile home or trailer", "other", "row or townhouse inner unit", "single family detached"),
                          to = c("other", "apartment or flat", "other", "other", "other", "other", "other", "other", "other", "other", "single family detached"))

# Print out a table of counts
table(expenditures$home.type)

```

**Problem 5:**

What fraction of the general population live in the Midwest, Northeast, South, and West regions of the US? What fraction of those in the 95% income percentile (i.e., the "95%") live in these regions? Do the percentages differ for the general population and the "95%"?

Create a data frame with the following columns:

1. `region`: Midwest, Northeast, South, West, and NA (not everyone reports their region!)
2. `general.pop`: the percentage of households in `expenditures` in each region
3. `rich.pop`: the percentage of households whose `income.rank` is greater than 0.95 in each region

(as a sanity check, both the `general.pop` and `rich.pop` columns should sum to 1)

Print this data frame as a table. Hint: you may want to use the function `merge()` to combine two data frames. (see the self-study material for lecture 2)

What are the main differences in the percentages for the general and rich population?


```{r}
# Find the proportion of the general population by region
proportions.by.region<- ddply(expenditures, "region", summarize, general.pop = length(housing)/number.of.records)

# Find the proportion of those in the top income ranks for each region
# Note that the proportion counts include those who did NOT report their income rank. This was a personal decision, and only affected the counts around 1%. 
the.ninety.five.percent = expenditures[!is.na(expenditures$income.rank) & expenditures$income.rank > .95,]
proportions.by.income<- ddply(the.ninety.five.percent, "region", summarize, rich.pop = length(income.rank)/nrow(the.ninety.five.percent))

# Merge the two data frames together
expenditures.proportions <- merge(proportions.by.region, proportions.by.income, all=TRUE)
kable(expenditures.proportions, digits = 3, col.names = c('Region', 'General Population', 'Rich Population'))
```


The south has the highest proportion of the population, and also the largest percentage of the wealthy. Interestingly, although the northeast has a small percentage of the population, almost 1/4 of the residents are wealthy. It was interesting to note that those who did not report their region did not have a single respondent falling in the upper 95%. This seemed unlikely, but validating with the raw data confirmed. 

As mentioned in the comments, I made the decision to include those who did NOT report their income rank in the proportion counts. 

